---
title: "P8105 Final Project"
output: github_document
---

```{r setup, include = FALSE}
library(tidyverse)
library(readxl)
library(mgcv)
library(modelr)
library(stringr)
library(patchwork)

knitr::opts_chunk$set(
  fig.width = 6,
  fig.asp = .6,
  out.width = "90%"
)
theme_set(theme_bw() + theme(legend.position = "bottom"))
options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_coulour_discrete = scale_color_viridis_d
scale_fill_discrete = scale_fill_viridis_d
```

## Data Import

```{r}
ghed_df <- 
  read_excel("data/GHED_data.XLSX")
```

### Safiya's section

__Indicators__
Primary health care - phc_usd_pc  
Infectious disease - dis1_usd2019  
Noncommunicable disease - dis4_usd2019  
Preventive care - hc6_usd2019  
Curative care - hc1_usd2019  
GDP - gdp_pc_usd  
CHE - che_pc_usd  

__Primary Health Care Spending__

Primary health care spending by income level (USD per capita)

```{r}
PHC_df <-
  ghed_df %>% 
    janitor::clean_names() %>% 
    filter(year == 2019) %>% 
    drop_na(phc_usd_pc) %>% 
    group_by(income_group) %>% 
    summarize(
      n_countries = n(),
      avg_phc = mean(phc_usd_pc))

PHC_df %>% 
  ggplot(aes(x = fct_relevel(as.factor(income_group), c("Low", "Low-Mid", "Up-Mid", "Hi")), y = avg_phc,  fill = income_group)) +
  geom_col() +
  labs(
    title = "Primary Health Care Spending",
    x = "Income level",
    y = "Average PHC spending (per capita USD)"
  ) + 
  theme(axis.text.x = element_text(angle = 90)) +
  theme(legend.position = "none")
```

Primary health care spending by income level (as as proportion of per capita GDP)

```{r}
PHC_df <-
  ghed_df %>% 
    janitor::clean_names() %>% 
    filter(year == 2019) %>% 
    drop_na(phc_usd_pc, gdp_pc_usd) %>% 
    group_by(income_group) %>% 
    summarize(
      n_countries = n(),
      avg_phc = mean(phc_usd_pc / gdp_pc_usd))

PHC_df %>% 
  ggplot(aes(x = fct_relevel(as.factor(income_group), c("Low", "Low-Mid", "Up-Mid", "Hi")), y = avg_phc,  fill = income_group)) +
  geom_col() +
  labs(
    title = "Primary Health Care Spending",
    x = "Income level",
    y = "Average PHC spending (as proportion of \n per capita GDP)"
  ) + 
  theme(axis.text.x = element_text(angle = 90)) +
  theme(legend.position = "none")
```

Primary health care spending by income level (as a proportion of per capita health expenditure)

```{r}
PHC_df <-
  ghed_df %>% 
    janitor::clean_names() %>% 
    filter(year == 2019) %>% 
    drop_na(phc_usd_pc, gdp_pc_usd) %>% 
    group_by(income_group) %>% 
    summarize(
      n_countries = n(),
      avg_phc = mean(phc_usd_pc / che_pc_usd))

PHC_df %>% 
  ggplot(aes(x = fct_relevel(as.factor(income_group), c("Low", "Low-Mid", "Up-Mid", "Hi")), y = avg_phc,  fill = income_group)) +
  geom_col() +
  labs(
    title = "Primary Health Care Spending",
    x = "Income level",
    y = "Average PHC spending (as proportion of \n per capita health expenditure)"
  ) + 
  theme(axis.text.x = element_text(angle = 90)) +
  theme(legend.position = "none")
```

__Preventive vs. Curative Spending__

Primary health care spending by income level (USD per capita)

```{r}
prev_df <-
  ghed_df %>% 
    janitor::clean_names() %>% 
    filter(year == 2019) %>% 
    drop_na(hc6_usd) %>% 
    group_by(income_group) %>% 
    summarize(
      n_countries = n(),
      `Preventive` = mean(hc6_usd2019))

cur_df <-
  ghed_df %>% 
    janitor::clean_names() %>% 
    filter(year == 2019) %>% 
    drop_na(hc1_usd) %>% 
    group_by(income_group) %>% 
    summarize(
      n_countries = n(),
      `Curative` = mean(hc1_usd2019))

prev_cur_df <-
  left_join(prev_df, cur_df) %>% 
  pivot_longer(
    `Preventive`:`Curative`,
    names_to = "Type of Spending",
    values_to = "avg_spend"
  )

prev_cur_df %>% 
  ggplot(aes(x = fct_relevel(as.factor(income_group), c("Low", "Low-Mid", "Up-Mid", "Hi")), y = avg_spend, fill = `Type of Spending`)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(
    title = "Preventive vs. Curative Spending",
    x = "Income level",
    y = "Average spending \n (million USD)"
  ) + 
  theme(axis.text.x = element_text(angle = 90))
```

Primary health care spending by income level (as as proportion of GDP)

```{r}
prev_df <-
  ghed_df %>% 
    janitor::clean_names() %>% 
    filter(year == 2019) %>% 
    drop_na(hc6_usd, gdp_pc_usd) %>% 
    group_by(income_group) %>% 
    summarize(
      n_countries = n(),
      `Preventive` = mean(hc6_usd2019 / gdp_usd2019))

cur_df <-
  ghed_df %>% 
    janitor::clean_names() %>% 
    filter(year == 2019) %>% 
    drop_na(hc1_usd, gdp_pc_usd) %>% 
    group_by(income_group) %>% 
    summarize(
      n_countries = n(),
      `Curative` = mean(hc1_usd2019 / gdp_usd2019))

prev_cur_df <-
  left_join(prev_df, cur_df) %>% 
  pivot_longer(
    `Preventive`:`Curative`,
    names_to = "Type of Spending",
    values_to = "avg_spend"
  )

prev_cur_df %>%
  ggplot(aes(x = fct_relevel(as.factor(income_group), c("Low", "Low-Mid", "Up-Mid", "Hi")), y = avg_spend, fill = `Type of Spending`)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(
    title = "Preventive vs. Curative Spending",
    x = "Income level",
    y = "Average spending (as proportion \n of GDP)"
  ) + 
  theme(axis.text.x = element_text(angle = 90))
```

Primary health care spending by income level (as proportion of health expenditure)

```{r}
prev_df <-
  ghed_df %>% 
    janitor::clean_names() %>% 
    filter(year == 2019) %>% 
    drop_na(hc6_usd, gdp_pc_usd) %>% 
    group_by(income_group) %>% 
    summarize(
      n_countries = n(),
      `Preventive` = mean(hc6_usd2019 / che_usd2019))

cur_df <-
  ghed_df %>% 
    janitor::clean_names() %>% 
    filter(year == 2019) %>% 
    drop_na(hc1_usd, gdp_pc_usd) %>% 
    group_by(income_group) %>% 
    summarize(
      n_countries = n(),
      `Curative` = mean(hc1_usd2019 / che_usd2019))

prev_cur_df <-
  left_join(prev_df, cur_df) %>% 
  pivot_longer(
    `Preventive`:`Curative`,
    names_to = "Type of Spending",
    values_to = "avg_spend"
  )

prev_cur_df %>%
  ggplot(aes(x = fct_relevel(as.factor(income_group), c("Low", "Low-Mid", "Up-Mid", "Hi")), y = avg_spend, fill = `Type of Spending`)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(
    title = "Preventive vs. Curative Spending",
    x = "Income level",
    y = "Average spending (as proportion \n of total health expenditure)"
  ) + 
  theme(axis.text.x = element_text(angle = 90))
```

__Infectious vs. Non-communicable Diseases__


### Ragyie's section

**Indicators for analysis**

* Curative care, in million constant (2019) US$ —> hc1_usd2019
* Preventive care, in million constant (2019) US$ —> hc6_usd2019
* Infectious and Parasitic Diseases, in million constant (2019) US$ —> dis1_usd2019
* Noncommunicable Diseases (NCDs), in million constant (2019) US$ —> dis4_usd2019
* Primary Health Care (PHC) Expenditure per Capita in US$ —> phc_usd_pc
* GDP - gdp_pc_usd  
* CHE - che_pc_usd 
* che_usd2019 
* gdp_usd2019

**Expenditure category analysis: WHO regions**

Will compare and plot the average in 2019 of spending in categories of interest for each WHO region 

**Primary health care**: 

Average primary health care expenditure for each WHO region in million constant USD$

```{r average_phc}
phc_ghed = 
  ghed_df %>% 
  janitor::clean_names() %>% 
  filter(year == 2019) %>% 
  drop_na(phc_usd_pc) %>% 
  group_by(region_who) %>% 
  summarize(
    n_countries = n(),
    avg_primary = mean(phc_usd_pc)
  )
```

Barplot: Primary health care expenditure in million constant USD$ 

```{r avg_phc_plot}
phc_ghed %>%   
  ggplot(aes(x = reorder(region_who, avg_primary), y = avg_primary, fill = region_who)) + 
  geom_bar(stat = "Identity") + 
  labs(
    title = "Primary health care spending by WHO region", 
    x = "WHO region", 
    y = "Average PHC spending in million constant USD$"
  ) + 
  theme(legend.position = "none")
```

Average primary health care expenditure for each WHO region with spending as a percent of GDP

To see how much different WHO regions are spending with regards to what resources they actually have

```{r average_phc_gdp}
phc_gdp_ghed =
  ghed_df %>% 
  janitor::clean_names() %>% 
  filter(year == 2019) %>% 
  drop_na(phc_usd_pc, gdp_pc_usd) %>% 
  group_by(region_who) %>% 
  summarize(
    n_countries = n(),
    avg_primary_gdp = mean(phc_usd_pc / gdp_pc_usd)
  )
```

Barplot: Primary health care expenditure with spending as a percent of GDP

```{r phc_gdp_plot}
phc_gdp_ghed %>%   
  ggplot(aes(x = reorder(region_who, avg_primary_gdp), y = avg_primary_gdp, fill = region_who)) + 
  geom_bar(stat = "Identity") + 
  labs(
    title = "Primary health care spending by WHO region", 
    x = "WHO region", 
    y = "Average PHC spending per capita GDP"
  ) + 
  theme(legend.position = "none")
```

Average primary health care expenditure for each WHO region with spending as a percentage of current health expenditure 

To investigate how much of health spending goes toward each category for each WHO region

```{r average_phc_che}
phc_che_ghed =
  ghed_df %>% 
  janitor::clean_names() %>% 
  filter(year == 2019) %>% 
  drop_na(phc_usd_pc, che_pc_usd) %>% 
  group_by(region_who) %>% 
  summarize(
    n_countries = n(),
    avg_primary_che = mean(phc_usd_pc / che_pc_usd)
  )
```

Barplot: Primary health care expenditure with spending as a percent of current health expenditure 

```{r phc_che_plot}
phc_che_ghed %>%   
  ggplot(aes(x = reorder(region_who, avg_primary_che), y = avg_primary_che, fill = region_who)) + 
  geom_bar(stat = "Identity") + 
  labs(
    title = "Primary health care spending by WHO region", 
    x = "WHO region", 
    y = "Average PHC spending per capita health expenditure"
  ) + 
  theme(legend.position = "none")
```

**Curative care vs. preventative care** 

Comparing average curative expenditure and average preventative expenditure for each WHO region

Average curative and preventative health care expenditure for each WHO region in million constant USD$

```{r avg_cp}
cp_df = 
  ghed_df %>% 
  janitor::clean_names() %>% 
  filter(year == 2019) %>% 
  drop_na(hc1_usd2019, hc6_usd2019) %>% 
  group_by(region_who) %>% 
  summarize(
    n_countries = n(),
    `Curative` = mean(hc1_usd2019),
    `Preventative` = mean(hc6_usd2019)
  ) %>% 
  pivot_longer(
    `Curative`:`Preventative`,
    names_to = "spending_type", 
    values_to = "average_spending"
  )
```

Barplot: Average curative vs. preventative health care expenditure for each WHO region in million constant USD$

```{r avg_cp_plot}
cp_df %>% 
  ggplot(aes(x = reorder(region_who, average_spending), y = average_spending, fill = spending_type)) + 
  geom_bar(stat = "Identity", position = "Dodge") + 
  labs(
    title = "Average curative vs. preventative spending by WHO region", 
    x = "WHO region", 
    y = "Average spending in million constant USD$"
  ) 
```

Average curative vs. preventative expenditure for each WHO region with spending as a percent of GDP

```{r avg_cp_gdp}
cp_gdp_df = 
  ghed_df %>% 
  janitor::clean_names() %>% 
  filter(year == 2019) %>% 
  drop_na(hc1_usd2019, hc6_usd2019, gdp_usd2019) %>% 
  group_by(region_who) %>% 
  summarize(
    n_countries = n(),
    `Curative` = mean(hc1_usd2019 / gdp_usd2019),
    `Preventative` = mean(hc6_usd2019 / gdp_usd2019)
  ) %>% 
  pivot_longer(
    `Curative`:`Preventative`,
    names_to = "spending_type", 
    values_to = "average_spending"
  )
```

Barplot: Curative vs. preventative expenditure with spending as a percent of GDP

```{r cp_gdp_plot}
cp_gdp_df %>% 
  ggplot(aes(x = reorder(region_who, average_spending), y = average_spending, fill = spending_type)) + 
  geom_bar(stat = "Identity", position = "Dodge") + 
  labs(
    title = "Average curative vs. preventative spending by WHO region", 
    x = "WHO region", 
    y = "Average spending per capita GDP"
  ) 
```

Average curative vs. preventative expenditure for each WHO region with spending as a percentage of current health expenditure

```{r avg_cp_che}
cp_che_df = 
  ghed_df %>% 
  janitor::clean_names() %>% 
  filter(year == 2019) %>% 
  drop_na(hc1_usd2019, hc6_usd2019, che_usd2019) %>% 
  group_by(region_who) %>% 
  summarize(
    n_countries = n(),
    `Curative` = mean(hc1_usd2019 / che_usd2019),
    `Preventative` = mean(hc6_usd2019 / che_usd2019)
  ) %>% 
  pivot_longer(
    `Curative`:`Preventative`,
    names_to = "spending_type", 
    values_to = "average_spending"
  )
```

Barplot: Curative vs. preventative expenditure with spending as a percent of current health expenditure 

```{r cp_che_plot}
cp_che_df %>% 
  ggplot(aes(x = reorder(region_who, average_spending), y = average_spending, fill = spending_type)) + 
  geom_bar(stat = "Identity", position = "Dodge") + 
  labs(
    title = "Average curative vs. preventative spending by WHO region", 
    x = "WHO region", 
    y = "Average spending per capita health expenditure"
  ) 
```

Combining our plots using patchwork to display on website 

**Primary health care expenditure **

per million constant USD$

```{r combined_phc_constant}
PHC_df <-
  ghed_df %>% 
    janitor::clean_names() %>% 
    filter(year == 2019) %>% 
    drop_na(phc_usd_pc) %>% 
    group_by(income_group) %>% 
    summarize(
      n_countries = n(),
      avg_phc = mean(phc_usd_pc))

income_constant = 
  PHC_df %>% 
  ggplot(aes(x = fct_relevel(as.factor(income_group), c("Low", "Low-Mid", "Up-Mid", "Hi")), y = avg_phc,  fill = income_group)) +
  geom_col() +
  labs(
    x = "Income level",
    y = "Average spending"
  ) + 
  theme(axis.text.x = element_text(angle = 90)) +
  theme(legend.position = "none")

phc_ghed = 
  ghed_df %>% 
  janitor::clean_names() %>% 
  filter(year == 2019) %>% 
  drop_na(phc_usd_pc) %>% 
  group_by(region_who) %>% 
  summarize(
    n_countries = n(),
    avg_primary = mean(phc_usd_pc)
  )

region_constant =
  phc_ghed %>%   
  ggplot(aes(x = reorder(region_who, avg_primary), y = avg_primary, fill = region_who)) + 
  geom_bar(stat = "Identity") + 
  labs(
    x = "WHO region", 
    y = "Average spending"
  ) + 
  theme(axis.text.x = element_text(angle = 90)) +
  theme(legend.position = "none")

combined_constant = income_constant + region_constant

combined_constant + 
  plot_annotation(
    title = "Primary health care spending in million constant USD$"
  )
```

per capita GDP

```{r combined_phc_gdp}
PHC_df =
  ghed_df %>% 
  janitor::clean_names() %>% 
  filter(year == 2019) %>% 
  drop_na(phc_usd_pc, gdp_pc_usd) %>% 
  group_by(income_group) %>% 
  summarize(
    n_countries = n(),
    avg_phc = mean(phc_usd_pc / gdp_pc_usd))

income_gdp = 
  PHC_df %>% 
  ggplot(aes(x = fct_relevel(as.factor(income_group), c("Low", "Low-Mid", "Up-Mid", "Hi")), y = avg_phc,  fill = income_group)) +
  geom_col() +
  labs(
    x = "Income level",
    y = "Average spending"
  ) + 
  theme(axis.text.x = element_text(angle = 90)) +
  theme(legend.position = "none")
  
phc_gdp_ghed =
  ghed_df %>% 
  janitor::clean_names() %>% 
  filter(year == 2019) %>% 
  drop_na(phc_usd_pc, gdp_pc_usd) %>% 
  group_by(region_who) %>% 
  summarize(
    n_countries = n(),
    avg_primary_gdp = mean(phc_usd_pc / gdp_pc_usd)
  )

region_gdp = 
  phc_gdp_ghed %>%   
  ggplot(aes(x = reorder(region_who, avg_primary_gdp), y = avg_primary_gdp, fill = region_who)) + 
  geom_bar(stat = "Identity") + 
  labs(
    x = "WHO region", 
    y = "Average spending"
  ) + 
  theme(axis.text.x = element_text(angle = 90)) +
  theme(legend.position = "none")

combined_gdp = income_gdp + region_gdp

combined_gdp + 
  plot_annotation(
    title = "Primary health care spending per capita GDP" 
  )
```



### Trisha's section

```{r}
ghed_df = 
  ghed_df %>%
  janitor::clean_names()
```

## Distribution of income levels of countries in each WHO region

```{r}
ghed_df %>%
  select(country, country_code, region_who, income_group) %>%
  distinct() %>%
  mutate(
    income_group = as.factor(income_group),
    income_group = factor(income_group, levels = c("Low","Low-Mid","Up-Mid","Hi"))
  ) %>%
  group_by(region_who) %>%
  mutate(
    num_countries = n()
  ) %>%
  group_by(region_who, income_group, num_countries) %>%
  summarize(n = n()) %>%
  mutate(
    perc_income_group = 100*(n/num_countries)
  ) %>%
  ggplot(aes(x=income_group,y=perc_income_group)) +
  geom_bar(aes(fill=income_group),stat="identity") +
  facet_grid(~region_who) +
  theme(axis.text.x = element_text(angle = 80, hjust = 1)) +
  ggtitle("Distribution of income levels of countries in each WHO region") +
  ylab("% of countries in region") +
  xlab("country income group (2019)")
```

The figure above shows the distribution of income levels of countries (categorized in 2019) in each WHO region: AFR (African Region), AMR (Region of the Americas), EMR (Eastern Mediterranean Region), EUR (European Region), SEAR (South-East Asian Region), and WPR (Western Pacific Region). \

Highest proportion of income group type by region: \
Low income: AFR \
Low-Mid income: SEAR \
Up-Mid income: AMR \
Hi income: EUR 


## Income group and total healthcare expenditure

```{r}
ghed_df %>%
  select(country, country_code, region_who, income_group, year, che_gdp, che_pc_usd) %>%
  filter(year == 2019) %>%
  mutate(
    income_group= as.factor(income_group),
    income_group = factor(income_group, levels = c("Low","Low-Mid","Up-Mid","Hi"))
  ) %>%
  group_by(income_group) %>%
  summarise(mean_che_gdp = mean(che_gdp,na.rm=TRUE), mean_che_pc_usd = mean(che_pc_usd,na.rm=TRUE)) %>%
  knitr::kable(digits = 2)
```

The above table describes the following variables for all income groups:\
che_gdp: Current Health Expenditure (CHE) as % Gross Domestic Product (GDP)	[Percentage]\
che_pc_usd: Current Health Expenditure (CHE) per Capita in US$ [Ones]\

As we can see, the mean CHE per capita in US$ increases as we go from low income groups to high income groups. The highest CHE as % of GDP belongs to countries in the highest income group at 7.67%; however, we see that this percentage for all income groups ranges from roughly 5-8%, so the amount per capita a country is able to dedicate for health expenditure depends and varies based on country GDP and income level.


## WHO region and total healthcare expenditure

```{r}
ghed_df %>%
  select(country, country_code, region_who, income_group, year, che_gdp, che_pc_usd) %>%
  filter(year == 2019) %>%
  group_by(region_who) %>%
  summarise(mean_che_gdp = mean(che_gdp,na.rm=TRUE), mean_che_pc_usd = mean(che_pc_usd,na.rm=TRUE)) %>%
  arrange(mean_che_pc_usd) %>%
  knitr::kable(digits = 2)
```
The above table describes the following variables for all WHO regions:\
che_gdp: Current Health Expenditure (CHE) as % Gross Domestic Product (GDP)	[Percentage]\
che_pc_usd: Current Health Expenditure (CHE) per Capita in US$ [Ones]\


## Developed countries - total spending over time 

major developed economies (UN): Canada, Japan, France, Germany, Italy, UK, US

```{r}
ghed_df %>%
  select(country, country_code, region_who, income_group, year, che_gdp, che_pc_usd) %>%
  filter(country == "United States of America" |
         country == "United Kingdom" |
         country == "Italy" |
         country == "Germany" |
         country == "France" |
         country == "Japan" |
         country == "Canada") %>%
  mutate(
    year = as.numeric(year)
  ) %>%
  group_by(country) %>%
  ggplot(aes(x=year,y=che_pc_usd, color=country)) +
  geom_line() +
  geom_point() +
  ggtitle("Total spending over time in developed countries") +
  scale_colour_viridis_d()
  
```

## Developed countries - government spending over time 

gghed_pc_usd Domestic General Government Health Expenditure (GGHE-D) per Capita in US$

```{r}
ghed_df %>%
  select(country, country_code, region_who, income_group, year, gghed_pc_usd) %>%
  filter(country == "United States of America" |
         country == "United Kingdom" |
         country == "Italy" |
         country == "Germany" |
         country == "France" |
         country == "Japan" |
         country == "Canada") %>%
  mutate(
    year = as.numeric(year)
  ) %>%
  group_by(country) %>%
  ggplot(aes(x=year,y=gghed_pc_usd, color=country)) +
  geom_line() +
  geom_point() +
  ggtitle("Government spending over time in developed countries") +
  scale_colour_viridis_d()
```


## Developing countries - total spending over time

developing economies: 
Morocco, Uganda, South Africa, Senegal
Nepal, Iraq, Haiti, Mexico, Argentina

```{r}
ghed_df %>%
  select(country, country_code, region_who, income_group, year, che_gdp, che_pc_usd) %>%
  filter(country == "Morocco" |
         country == "Uganda" |
         country == "South Africa" |
         country == "Senegal" |
         country == "Nepal" |
         country == "Iraq" |
         country == "Haiti" |
         country == "Mexico" |
         country == "Argentina") %>%
  mutate(
    year = as.numeric(year)
  ) %>%
  group_by(country) %>%
  ggplot(aes(x=year,y=che_pc_usd, color=country)) +
  geom_line() +
  geom_point() +
  ggtitle("Total spending over time in developing countries") +
  scale_colour_viridis_d()
```

## Developing countries - government spending over time

```{r}
ghed_df %>%
  select(country, country_code, region_who, income_group, year, gghed_pc_usd) %>%
  filter(country == "Morocco" |
         country == "Uganda" |
         country == "South Africa" |
         country == "Senegal" |
         country == "Nepal" |
         country == "Iraq" |
         country == "Haiti" |
         country == "Mexico" |
         country == "Argentina") %>%
  mutate(
    year = as.numeric(year)
  ) %>%
  group_by(country) %>%
  ggplot(aes(x=year,y=gghed_pc_usd, color=country)) +
  geom_line() +
  geom_point() +
  ggtitle("Government spending over time in developing countries") +
  scale_colour_viridis_d()
```


## LM - correlation between WHO region, external funding, private spending, and government expenditure and income group

```{r}

ghed_lm_df = 
  ghed_df %>%
  select(country, country_code, region_who, income_group, ext_pc_usd, pvtd_pc_usd, gghed_pc_usd) %>%
  drop_na() %>%
  mutate(
    income_group = as.factor(income_group),
    income_group = factor(income_group, levels = c("Low","Low-Mid","Up-Mid","Hi")),
    income_group = fct_recode(income_group, '1'="Low",'2'="Low-Mid", '3'="Up-Mid", '4'="Hi")
  )

```

```{r}
ghed_lm_df_shuffled = ghed_lm_df[sample(nrow(ghed_lm_df)),]

ghed_lm_df_shuffled = 
  ghed_lm_df_shuffled %>%
  mutate(
    sum_pc = ext_pc_usd + pvtd_pc_usd + gghed_pc_usd,
    ext_pc_usd = ext_pc_usd/sum_pc,
    pvtd_pc_usd = pvtd_pc_usd/sum_pc,
    gghed_pc_usd = gghed_pc_usd/sum_pc
  )
  
```

```{r}
fit_exp_dstbn = lm(as.numeric(income_group) ~ ext_pc_usd + pvtd_pc_usd + gghed_pc_usd, ghed_lm_df_shuffled)
fit_region_exp = lm(as.numeric(income_group) ~ ext_pc_usd + pvtd_pc_usd + gghed_pc_usd + region_who, ghed_lm_df_shuffled)

summary(fit_exp_dstbn)

fit_region_exp %>% 
  broom::glance()
```




### Pooja's Part 

```{r}
#Developed - Canada, Japan, France, Germany, Italy, UK, US

ghed_df%>%
  filter(country=="Canada"| country=="Japan"| country=="France"| country=="Germany"| country=="Italy"| country=="United Kingdom
"| country=="United States of America")%>%
  filter(year == 2019)%>%
  select(country, year, gghed_che, pvtd_che)%>%
  mutate(sum_per = gghed_che + pvtd_che)%>%
  pivot_longer(cols = contains("che"),
               names_to = "exp_type",
               values_to = "exp_per")%>%
  ggplot(aes(fill = exp_type, y = exp_per, x = country))+
  geom_bar(position="stack", stat="identity")
```

```{r}
ghed_df%>%
  filter(year == 2019)%>%
  select(region_who, income_group, year, gghed_che, pvtd_che)%>%
  group_by(region_who, income_group)%>%
  mutate(sum_gghed = sum(gghed_che),
         sum_pvtd = sum(pvtd_che), 
         sum_che = sum_gghed + sum_pvtd,
         income_group = factor(income_group, levels = c("Low", "Low-Mid", "Up-Mid", "Hi")))%>%
  select(region_who, income_group, year, sum_gghed, sum_pvtd, sum_che)%>%
  distinct()%>%
  mutate(gghed_per = sum_gghed/sum_che, 
         pvtd_per = sum_pvtd/sum_che)%>%
  pivot_longer(cols = contains("per"),
               names_to = "exp_type",
               values_to = "exp_per")%>%
  select(region_who, income_group, exp_type, exp_per)%>%
  ggplot(aes(fill = exp_type, y = exp_per, x = income_group))+
  geom_bar(stat="identity")+facet_grid(cols = vars(region_who))+theme(axis.text.x = element_text(angle = 45, vjust = 0.5, hjust=1))+
scale_fill_discrete(name="Expenditure Type",
                         breaks=c("gghed_per", "pvtd_per"),
                         labels=c("Government", "Private"))

```

Government vs. Private expenditure over time in the US 
```{r}
ghed_df%>%
  filter(country == "United States of America")%>%
  select(country, year, gghed_pc_usd, pvtd_pc_usd)%>%
  pivot_longer(cols = contains("pc_usd"),
               names_to = "exp_type", 
               values_to = "exp_per_cap")%>%
  ggplot() + geom_point(aes(x = year, y = exp_per_cap, col = exp_type))+geom_line(aes(x = year, y = exp_per_cap, group = exp_type, colour = exp_type))
  


```

```{r}
ghed_df%>%
  filter(country=="United States of America")%>%
  select(country, year, gghed_che, pvtd_che)%>%
  mutate(sum_per = gghed_che + pvtd_che)%>%
  pivot_longer(cols = contains("che"),
               names_to = "exp_type",
               values_to = "exp_per")%>%
  ggplot() + geom_point(aes(x = year, y = exp_per, col = exp_type))+geom_line(aes(x = year, y = exp_per, group = exp_type, colour = exp_type))
```

```{r}
ghed_df%>%
  filter(year ==2019)%>%
  select(country,region_who, year,income_group, gghed_pc_usd, pvtd_pc_usd, ext_pc_usd)%>%
  group_by(region_who, income_group)%>%
  mutate(avg_ext_exp = mean(ext_pc_usd))%>%
  select(region_who, income_group, avg_ext_exp)%>%
  distinct()%>%
    ggplot(aes(y = avg_ext_exp, x = income_group))+
  geom_bar(stat="identity")+facet_grid(cols = vars(region_who))+theme(axis.text.x = element_text(angle = 45, vjust = 0.5, hjust=1))
```

```{r}
ghed_df%>%
  filter(year ==2019)%>%
  select(country,region_who, year,income_group, gghed_pc_usd, pvtd_pc_usd, ext_che)%>%
  group_by(region_who, income_group)%>%
  mutate(avg_ext_exp = mean(ext_che), 
         income_group = factor(income_group, levels = c("Low", "Low-Mid", "Up-Mid", "Hi")))%>%
  select(region_who, income_group, avg_ext_exp)%>%
  distinct()%>%
    ggplot(aes(y = avg_ext_exp, x = income_group))+
  geom_bar(stat="identity")+facet_grid(cols = vars(region_who))+theme(axis.text.x = element_text(angle = 45, vjust = 0.5, hjust=1))
```

```{r}
un_country_class<- read_csv("data/un_country_classification.csv")

head(un_country_class)

un_country_class<- un_country_class%>%
  janitor::clean_names()%>%
  select(country_or_area, developed_developing_countries)%>%
  mutate(developed = str_replace(developed_developing_countries, ".*Developing.*", "Developing"))%>%
  select(country_or_area, developed)



ghed_df<- left_join(ghed_df, un_country_class, by = c("country" = "country_or_area"))

```

```{r}
 ghed_df%>%
  select(country,region_who, year,developed, gghed_pc_usd, pvtd_pc_usd, ext_che)%>%
  group_by(region_who, developed)%>%
  mutate(avg_ext_exp = mean(ext_che,na.rm=TRUE), 
         developed = factor(developed, levels = c(NA, "Developing", "Developed")))%>%
  select(region_who, developed, avg_ext_exp)%>%
  distinct()%>%
    ggplot(aes(y = avg_ext_exp, x = developed, colours=region_who))+
  geom_bar(stat="identity", aes(fill = region_who))+facet_grid(cols = vars(region_who))+theme(axis.text.x = element_text(angle = 45, vjust = 0.5, hjust=1))+scale_fill_viridis_d()
```
