---
title: "P8105 Final Project"
output: github_document
---

```{r setup, include = FALSE}
library(tidyverse)
library(readxl)

knitr::opts_chunk$set(
  fig.width = 6,
  fig.asp = .6,
  out.width = "90%"
)
theme_set(theme_bw() + theme(legend.position = "bottom"))
options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_coulour_discrete = scale_color_viridis_d
scale_fill_discrete = scale_fill_viridis_d
```

## Data Import

```{r}
ghed_df <- 
  read_excel("data/GHED_data.XLSX")
```

### Safiya's section

__Indicators__
Primary health care - phc_usd_pc  
Infectious disease - dis1_usd2019  
Noncommunicable disease - dis4_usd2019  
Preventive care - hc6_usd2019  
Curative care - hc1_usd2019  
GDP - gdp_pc_usd  
CHE - che_pc_usd  

__Primary Health Care Spending__

Primary health care spending by income level (USD per capita)

```{r}
PHC_df <-
  ghed_df %>% 
    janitor::clean_names() %>% 
    filter(year == 2019) %>% 
    drop_na(phc_usd_pc) %>% 
    group_by(income_group) %>% 
    summarize(
      n_countries = n(),
      avg_phc = mean(phc_usd_pc))

PHC_df %>% 
  ggplot(aes(x = fct_relevel(as.factor(income_group), c("Low", "Low-Mid", "Up-Mid", "Hi")), y = avg_phc,  fill = income_group)) +
  geom_col() +
  labs(
    title = "Primary Health Care Spending",
    x = "Income level",
    y = "Average PHC spending (per capita USD)"
  ) + 
  theme(axis.text.x = element_text(angle = 90)) +
  theme(legend.position = "none")
```

Primary health care spending by income level (as as proportion of per capita GDP)

```{r}
PHC_df <-
  ghed_df %>% 
    janitor::clean_names() %>% 
    filter(year == 2019) %>% 
    drop_na(phc_usd_pc, gdp_pc_usd) %>% 
    group_by(income_group) %>% 
    summarize(
      n_countries = n(),
      avg_phc = mean(phc_usd_pc / gdp_pc_usd))

PHC_df %>% 
  ggplot(aes(x = fct_relevel(as.factor(income_group), c("Low", "Low-Mid", "Up-Mid", "Hi")), y = avg_phc,  fill = income_group)) +
  geom_col() +
  labs(
    title = "Primary Health Care Spending",
    x = "Income level",
    y = "Average PHC spending (as proportion of \n per capita GDP)"
  ) + 
  theme(axis.text.x = element_text(angle = 90)) +
  theme(legend.position = "none")
```

Primary health care spending by income level (as a proportion of per capita health expenditure)

```{r}
PHC_df <-
  ghed_df %>% 
    janitor::clean_names() %>% 
    filter(year == 2019) %>% 
    drop_na(phc_usd_pc, gdp_pc_usd) %>% 
    group_by(income_group) %>% 
    summarize(
      n_countries = n(),
      avg_phc = mean(phc_usd_pc / che_pc_usd))

PHC_df %>% 
  ggplot(aes(x = fct_relevel(as.factor(income_group), c("Low", "Low-Mid", "Up-Mid", "Hi")), y = avg_phc,  fill = income_group)) +
  geom_col() +
  labs(
    title = "Primary Health Care Spending",
    x = "Income level",
    y = "Average PHC spending (as proportion of \n per capita health expenditure)"
  ) + 
  theme(axis.text.x = element_text(angle = 90)) +
  theme(legend.position = "none")
```

__Preventive vs. Curative Spending__

Primary health care spending by income level (USD per capita)

```{r}
prev_df <-
  ghed_df %>% 
    janitor::clean_names() %>% 
    filter(year == 2019) %>% 
    drop_na(hc6_usd) %>% 
    group_by(income_group) %>% 
    summarize(
      n_countries = n(),
      `Preventive` = mean(hc6_usd2019))

cur_df <-
  ghed_df %>% 
    janitor::clean_names() %>% 
    filter(year == 2019) %>% 
    drop_na(hc1_usd) %>% 
    group_by(income_group) %>% 
    summarize(
      n_countries = n(),
      `Curative` = mean(hc1_usd2019))

prev_cur_df <-
  left_join(prev_df, cur_df) %>% 
  pivot_longer(
    `Preventive`:`Curative`,
    names_to = "Type of Spending",
    values_to = "avg_spend"
  )

prev_cur_df %>% 
  ggplot(aes(x = fct_relevel(as.factor(income_group), c("Low", "Low-Mid", "Up-Mid", "Hi")), y = avg_spend, fill = `Type of Spending`)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(
    title = "Preventive vs. Curative Spending",
    x = "Income level",
    y = "Average spending \n (million USD)"
  ) + 
  theme(axis.text.x = element_text(angle = 90))
```

Primary health care spending by income level (as as proportion of GDP)

```{r}
prev_df <-
  ghed_df %>% 
    janitor::clean_names() %>% 
    filter(year == 2019) %>% 
    drop_na(hc6_usd, gdp_pc_usd) %>% 
    group_by(income_group) %>% 
    summarize(
      n_countries = n(),
      `Preventive` = mean(hc6_usd2019 / gdp_usd2019))

cur_df <-
  ghed_df %>% 
    janitor::clean_names() %>% 
    filter(year == 2019) %>% 
    drop_na(hc1_usd, gdp_pc_usd) %>% 
    group_by(income_group) %>% 
    summarize(
      n_countries = n(),
      `Curative` = mean(hc1_usd2019 / gdp_usd2019))

prev_cur_df <-
  left_join(prev_df, cur_df) %>% 
  pivot_longer(
    `Preventive`:`Curative`,
    names_to = "Type of Spending",
    values_to = "avg_spend"
  )

prev_cur_df %>%
  ggplot(aes(x = fct_relevel(as.factor(income_group), c("Low", "Low-Mid", "Up-Mid", "Hi")), y = avg_spend, fill = `Type of Spending`)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(
    title = "Preventive vs. Curative Spending",
    x = "Income level",
    y = "Average spending (as proportion \n of GDP)"
  ) + 
  theme(axis.text.x = element_text(angle = 90))
```

Primary health care spending by income level (as proportion of health expenditure)

```{r}
prev_df <-
  ghed_df %>% 
    janitor::clean_names() %>% 
    filter(year == 2019) %>% 
    drop_na(hc6_usd, gdp_pc_usd) %>% 
    group_by(income_group) %>% 
    summarize(
      n_countries = n(),
      `Preventive` = mean(hc6_usd2019 / che_usd2019))

cur_df <-
  ghed_df %>% 
    janitor::clean_names() %>% 
    filter(year == 2019) %>% 
    drop_na(hc1_usd, gdp_pc_usd) %>% 
    group_by(income_group) %>% 
    summarize(
      n_countries = n(),
      `Curative` = mean(hc1_usd2019 / che_usd2019))

prev_cur_df <-
  left_join(prev_df, cur_df) %>% 
  pivot_longer(
    `Preventive`:`Curative`,
    names_to = "Type of Spending",
    values_to = "avg_spend"
  )

prev_cur_df %>%
  ggplot(aes(x = fct_relevel(as.factor(income_group), c("Low", "Low-Mid", "Up-Mid", "Hi")), y = avg_spend, fill = `Type of Spending`)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(
    title = "Preventive vs. Curative Spending",
    x = "Income level",
    y = "Average spending (as proportion \n of total health expenditure)"
  ) + 
  theme(axis.text.x = element_text(angle = 90))
```

__Infectious vs. Non-communicable Diseases__


### Ragyie's section

**Indicators for analysis**

* Curative care, in million constant (2019) US$ —> hc1_usd2019
* Preventive care, in million constant (2018) US$ —> hc6_usd2019
* Infectious and Parasitic Diseases, in million constant (2018) US$ —> dis1_usd2019
* Noncommunicable Diseases (NCDs), in million constant (2018) US$ —> dis4_usd2019
* Primary Health Care (PHC) Expenditure per Capita in US$ —> phc_usd_pc
* GDP - gdp_pc_usd  
* CHE - che_pc_usd 
* che_usd_2019 
* gdp_usd_2019

**Expenditure category analysis: WHO regions**

Will compare and plot the average in 2019 of spending in categories of interest for each WHO region 

**Primary health care**: 

Average primary health care expenditure for each WHO region in million constant USD$

```{r average_phc}
phc_ghed = 
  ghed_df %>% 
  janitor::clean_names() %>% 
  filter(year == 2019) %>% 
  drop_na(phc_usd_pc) %>% 
  group_by(region_who) %>% 
  summarize(
    n_countries = n(),
    avg_primary = mean(phc_usd_pc)
  )
```

Barplot: Primary health care expenditure in million constant USD$ 

```{r avg_phc_plot}
phc_ghed %>%   
  ggplot(aes(x = reorder(region_who, avg_primary), y = avg_primary, fill = region_who)) + 
  geom_bar(stat = "Identity") + 
  labs(
    title = "Primary health care spending", 
    x = "WHO region", 
    y = "Average PHC spending in million constant USD$"
  ) + 
  theme(legend.position = "none")
```


